---
title: "GGRC32_project"
author: "Zehui Yin"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r setup}
library(tidyverse)
library(jsonlite)
library(arrow)
library(sf)
library(sp)
library(mapview)
library(sfheaders)
library(stringr)
Sys.setenv(JAVA_HOME="C:\\Program Files\\Java\\jdk-11.0.17")
library(r5r)
```

Read and combine trips data

```{r}
trips1 <- read.csv("./Data/Bike share ridership 2021-01.csv")
trips2 <- read.csv("./Data/Bike share ridership 2021-02.csv")
trips3 <- read.csv("./Data/Bike share ridership 2021-03.csv")
trips4 <- read.csv("./Data/Bike share ridership 2021-04.csv")
trips5 <- read.csv("./Data/Bike share ridership 2021-05.csv")
trips6 <- read.csv("./Data/Bike share ridership 2021-06.csv")
trips7 <- read.csv("./Data/Bike share ridership 2021-07.csv")
trips8 <- read.csv("./Data/Bike share ridership 2021-08.csv")
trips9 <- read.csv("./Data/Bike share ridership 2021-09.csv")
trips10 <- read.csv("./Data/Bike share ridership 2021-10.csv")
trips11 <- read.csv("./Data/Bike share ridership 2021-11.csv")
trips12 <- read.csv("./Data/Bike share ridership 2021-12.csv")
trips <- rbind(trips1, trips2, trips3, trips4, trips5, trips6, trips7, trips8, trips9, trips10, trips11, trips12)
trips$Start.Time <- as.POSIXct(trips$Start.Time, "%m/%d/%Y %H:%M", tz = "America/Toronto")
trips$End.Time <- as.POSIXct(trips$End.Time, "%m/%d/%Y %H:%M", tz = "America/Toronto")

write_parquet(trips, "./Data/trips.parquet")
trips <- read_parquet("./Data/trips.parquet") # only run this line is fine
```

Read and process station data

```{r}
stations <- read_json("./Data/station_information.json")
stations <- stations$data
stations <- stations$stations
stations_df <- as.data.frame(matrix(nrow = length(stations), ncol = 14))
colnames(stations_df) <- c("station_id", "name", "physical_configuration",
                           "lat", "lon", "altitude", "address", "capacity",
                           "is_charging_station", "rental_methods", "groups",
                           "obcn", "nearby_distance", "ride_code_support")
for (i in 1:length(stations)) {
  holder <- as.data.frame(unlist(stations[i]))
  stations_df[i, "station_id"] <- holder["station_id",]
  stations_df[i, "name"] <- holder["name",]
  stations_df[i, "physical_configuration"] <- holder["physical_configuration",]
  stations_df[i, "lat"] <- holder["lat",]
  stations_df[i, "lon"] <- holder["lon",]
  stations_df[i, "altitude"] <- holder["altitude",]
  stations_df[i, "address"] <- holder["address",]
  stations_df[i, "capacity"] <- holder["capacity",]
  stations_df[i, "is_charging_station"] <- holder["is_charging_station",]
  stations_df[i, "rental_methods"] <- paste(holder[str_which(row.names(holder), coll("rental_methods", ignore_case = FALSE, locale = "en")),], collapse = ",")
  stations_df[i, "groups"] <- holder["groups",]
  stations_df[i, "obcn"] <- holder["obcn",]
  stations_df[i, "nearby_distance"] <- holder["nearby_distance",]
  stations_df[i, "ride_code_support"] <- holder["_ride_code_support",]
}

write_parquet(stations_df, "./Data/stations.parquet")
stations_df <- read_parquet("./Data/stations.parquet") # only run this line is fine
```

Routing setup

```{r}
options(java.parameters = "-Xmx12G")
r5r_core <- setup_r5(data_path = "./Network", verbose = F)
```

# Travel time matrix between all possible combinations of stations

```{r}
stations <- stations_df[, c("station_id", "lat", "lon")]
stations <- st_as_sf(stations, coords = c("lon", "lat"), crs = 4326, agr = "constant")
# mapview(stations)

origins <- stations[stations$station_id %in% unique(trips$Start.Station.Id),]
names(origins)[1] <- "id"
destinations <- stations[stations$station_id %in% unique(trips$End.Station.Id),]
names(destinations)[1] <- "id"

mode <- "BICYCLE"
max_walk_dist <- 1000L
max_trip_duration <- 1000L
bike_speed <- 12L
max_lts <- 1L
departure_datetime <- as.POSIXct("1-1-2023 14:00:00", tz = "America/Toronto",
                                 format = "%d-%m-%Y %H:%M:%S")

# Safest
max_lts <- 1L
ttm_lts1 <- travel_time_matrix(r5r_core = r5r_core,
                          origins = origins,
                          destinations = destinations,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          bike_speed = bike_speed,
                          max_lts = max_lts,
                          verbose = FALSE)

# normal
max_lts <- 2L
ttm_lts2 <- travel_time_matrix(r5r_core = r5r_core,
                          origins = origins,
                          destinations = destinations,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          bike_speed = bike_speed,
                          max_lts = max_lts,
                          verbose = FALSE)

# danger
max_lts <- 3L
ttm_lts3 <- travel_time_matrix(r5r_core = r5r_core,
                          origins = origins,
                          destinations = destinations,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          bike_speed = bike_speed,
                          max_lts = max_lts,
                          verbose = FALSE)

# dangerest
max_lts <- 4L
ttm_lts4 <- travel_time_matrix(r5r_core = r5r_core,
                          origins = origins,
                          destinations = destinations,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          bike_speed = bike_speed,
                          max_lts = max_lts,
                          verbose = FALSE)
```

Visualize the travel time distribution

```{r}
par(mfrow = c(2,2))
hist(ttm_lts1$travel_time, breaks = seq(0, max(ttm_lts1$travel_time) + 10, 10), ylim = c(0, 70000))
hist(ttm_lts2$travel_time, breaks = seq(0, max(ttm_lts1$travel_time) + 10, 10), ylim = c(0, 70000))
hist(ttm_lts3$travel_time, breaks = seq(0, max(ttm_lts1$travel_time) + 10, 10), ylim = c(0, 70000))
hist(ttm_lts4$travel_time, breaks = seq(0, max(ttm_lts1$travel_time) + 10, 10), ylim = c(0, 70000))
```

## Connectivity at stations

Reweighting based on trip count at actual origin and destination
Assignment more weights to OD pairs occurs more in the dataset

$$C_i = \frac{\sum_{j=1}^Nn_{ij}*t_{ij}}{\sum_{j=1}^Nn_{ij}}$$

Where:
+ $C_i$ is the connectivity score or total travel time originated from $i^{th}$ station in 2021
+ $N$ is the total number of stations
+ $n_{ij}$ is the total number of trips originated from $i^{th}$ station to $j^{th}$ station
+ $t_{ij}$ is the travel time from $i^{th}$ station to $j^{th}$ station

```{r}
stations_connectivity <- origins
stations_connectivity <- as.data.frame(stations_connectivity)
stations_connectivity$connectivity_lts1 <- NA
stations_connectivity$connectivity_lts2 <- NA
stations_connectivity$connectivity_lts3 <- NA
stations_connectivity$connectivity_lts4 <- NA

# extreme long time to process
#N <- nrow(stations_connectivity)
#for (i in 1:N) {
#  denominator <- 
#    nrow(trips[which(trips$Start.Station.Id == stations_connectivity$id[i]),])
#  numerator <- 0
#  for (j in 1:N) {
#    numerator <- numerator +
#      nrow(trips[which(trips$Start.Station.Id == stations_connectivity$id[i] & 
#                       trips$End.Station.Id == stations_connectivity$id[j]),]) *
#      ttm_lts1[which(ttm_lts1$fromId == stations_connectivity$id[i] & 
#                     ttm_lts1$toId == stations_connectivity$id[j]),
#               "travel_time"]
#  }
#}

# improve code efficiency
#N <- nrow(stations_connectivity)
#for (i in 1:N) {
#  trips_i <- 
#    trips[which(trips$Start.Station.Id == stations_connectivity$id[i]),]
#  denominator <- nrow(trips_i)
#  ttm_lts1_i <- 
#    ttm_lts1[which(ttm_lts1$fromId == stations_connectivity$id[i]),] 
#  numerator <- 0
#  for (j in 1:N) {
#    toid <- stations_connectivity$id[j]
#    numerator <- numerator +
#      nrow(trips_i[which(trips_i$End.Station.Id == toid),]) *
#      ttm_lts1_i[which(ttm_lts1_i$toId == toid), "travel_time"]
#  }
#}

# not working for some reasons
connectivity <- function(trips, ttm, output) {
  N <- nrow(output)
  total_loop_number <- N*N
  for (i in 1:N) {
    trips_i <- 
      trips[which(trips$Start.Station.Id == output$id[i]),]
    denominator <- nrow(trips_i)
    ttm_i <- 
      ttm[which(ttm$fromId == output$id[i]),]
    numerator <- 0
    for (j in 1:N) {
      toid <- output$id[j]
      numerator <- numerator +
        nrow(trips_i[which(trips_i$End.Station.Id == toid),]) *
        ttm_i[which(ttm_i$toId == toid), "travel_time"]
    }
    output[i, paste("connectivity", substr(deparse(quote(ttm)),4,8))] <-
      numerator/denominator
    cat("Process completed", i*N/total_loop_number*100, "%\n")
  }
}

connectivity(trips, ttm_lts1, stations_connectivity)
```

# Shortest path between each stations

```{r}
stations <- stations_df[, c("name", "lat", "lon")]
stations <- st_as_sf(stations, coords = c("lon", "lat"), crs = 4326, agr = "constant")
# mapview(stations)

origins <- stations[stations$name %in% unique(trips$Start.Station.Name),]
names(origins)[1] <- "id"
origins <- origins[rep(seq_len(nrow(origins)), each = 577), ]

destinations <- stations[stations$name %in% unique(trips$End.Station.Name),]
names(destinations)[1] <- "id"
destinations <- destinations[rep(1:577, 577), ]

mode <- "BICYCLE"
max_walk_dist <- 1000L
max_trip_duration <- 1000L
bike_speed <- 12L
max_lts <- 2L
departure_datetime <- as.POSIXct("1-4-2021 14:00:00", tz = "America/Toronto",
                                 format = "%d-%m-%Y %H:%M:%S")

routes <- detailed_itineraries(r5r_core = r5r_core,
                          origins = origins,
                          destinations = destinations,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          bike_speed = bike_speed,
                          max_lts = max_lts,
                          verbose = FALSE)

# remove start and end at the same place
routes <- routes[which(routes$fromId != routes$toId),]
```

## How fast does riders travel within the Toronto Bikeshare System at stations level (also seasonal variation)

```{r}
stations_used <- 
  stations_df[stations_df$name %in% unique(trips$Start.Station.Name) | 
             stations_df$name %in% unique(trips$End.Station.Name), ]

trips <- trips[trips$Start.Station.Name %in% stations_used$name & 
                 trips$End.Station.Name %in% stations_used$name,]

# summary(as.factor(months(trips$Start.Time)))
trips_spring <- trips[months(trips$Start.Time) %in% c("March", "April", "May"),] # 805719
trips_summer <- trips[months(trips$Start.Time) %in% c("June", "July", "August"),] # 1413500
trips_autumn <- trips[months(trips$Start.Time) %in% c("September", "October", "November"),] # 1064864
trips_winter <- trips[months(trips$Start.Time) %in% c("December", "January", "February"),] # 291099
```

$$S_i = \frac{\sum_{j=1}^{M-1}n_{ij}*d_{ij}}{\sum_{z=1}^{N_i}t_{iz}}$$

where:
+ $S_i$: mean travel speed of trips originated from $i^{th}$ station (m/s)
+ $M$: number of total stations
+ $N_i$: total number of trips originated from $i^{th}$ station
+ $n_{ij}$: number of trips originated from $i^{th}$ station to $j^{th}$ station
+ $d_{ij}$: distance travelled from $i^{th}$ station to $j^{th}$ station
+ $t_{iz}$: time it took by $z^{th}$ trip to travel from $i^{th}$ station

note: trips start and end at the same station are omitted in the calculation

```{r}
routes_df <- as.data.frame(routes)

# all seasons
for (i in stations_used$name) {
  trips_starts_at_i <- 
    trips[which(trips$Start.Station.Name == i & trips$End.Station.Name != i), ]
  total_time <- sum(as.numeric(trips_starts_at_i[, "Trip..Duration"]))
  distance_matrix <- 
    routes_df[which(routes_df$fromId == i), c("toId", "distance")]
  total_distance <- 0
  for (j in distance_matrix$toId) {
    total_distance <- total_distance +
      nrow(trips_starts_at_i[trips_starts_at_i$End.Station.Name == j,]) *
      distance_matrix[distance_matrix$toId == j, "distance"]
  }
  stations_used[stations_used$name == i, "all_mean_speed_ms"] <-
    total_distance/total_time
}
stations_used$all_mean_speed_kmh <- stations_used$all_mean_speed_ms * 3.6

# spring
for (i in stations_used$name) {
  trips_starts_at_i <- 
    trips_spring[which(trips_spring$Start.Station.Name == i & trips_spring$End.Station.Name != i), ]
  total_time <- sum(as.numeric(trips_starts_at_i[, "Trip..Duration"]))
  distance_matrix <- 
    routes_df[which(routes_df$fromId == i), c("toId", "distance")]
  total_distance <- 0
  for (j in distance_matrix$toId) {
    total_distance <- total_distance +
      nrow(trips_starts_at_i[trips_starts_at_i$End.Station.Name == j,]) *
      distance_matrix[distance_matrix$toId == j, "distance"]
  }
  stations_used[stations_used$name == i, "spring_mean_speed_ms"] <-
    total_distance/total_time
}
stations_used$spring_mean_speed_kmh <- stations_used$spring_mean_speed_ms * 3.6

# summer
for (i in stations_used$name) {
  trips_starts_at_i <- 
    trips_summer[which(trips_summer$Start.Station.Name == i & trips_summer$End.Station.Name != i), ]
  total_time <- sum(as.numeric(trips_starts_at_i[, "Trip..Duration"]))
  distance_matrix <- 
    routes_df[which(routes_df$fromId == i), c("toId", "distance")]
  total_distance <- 0
  for (j in distance_matrix$toId) {
    total_distance <- total_distance +
      nrow(trips_starts_at_i[trips_starts_at_i$End.Station.Name == j,]) *
      distance_matrix[distance_matrix$toId == j, "distance"]
  }
  stations_used[stations_used$name == i, "summer_mean_speed_ms"] <-
    total_distance/total_time
}
stations_used$summer_mean_speed_kmh <- stations_used$summer_mean_speed_ms * 3.6

# autumn
for (i in stations_used$name) {
  trips_starts_at_i <- 
    trips_autumn[which(trips_autumn$Start.Station.Name == i & trips_autumn$End.Station.Name != i), ]
  total_time <- sum(as.numeric(trips_starts_at_i[, "Trip..Duration"]))
  distance_matrix <- 
    routes_df[which(routes_df$fromId == i), c("toId", "distance")]
  total_distance <- 0
  for (j in distance_matrix$toId) {
    total_distance <- total_distance +
      nrow(trips_starts_at_i[trips_starts_at_i$End.Station.Name == j,]) *
      distance_matrix[distance_matrix$toId == j, "distance"]
  }
  stations_used[stations_used$name == i, "autumn_mean_speed_ms"] <-
    total_distance/total_time
}
stations_used$autumn_mean_speed_kmh <- stations_used$autumn_mean_speed_ms * 3.6

# winter
for (i in stations_used$name) {
  trips_starts_at_i <- 
    trips_winter[which(trips_winter$Start.Station.Name == i & trips_winter$End.Station.Name != i), ]
  total_time <- sum(as.numeric(trips_starts_at_i[, "Trip..Duration"]))
  distance_matrix <- 
    routes_df[which(routes_df$fromId == i), c("toId", "distance")]
  total_distance <- 0
  for (j in distance_matrix$toId) {
    total_distance <- total_distance +
      nrow(trips_starts_at_i[trips_starts_at_i$End.Station.Name == j,]) *
      distance_matrix[distance_matrix$toId == j, "distance"]
  }
  stations_used[stations_used$name == i, "winter_mean_speed_ms"] <-
    total_distance/total_time
}
stations_used$winter_mean_speed_kmh <- stations_used$winter_mean_speed_ms * 3.6

stations_used$seasonal_variation <- sqrt(((stations_used$spring_mean_speed_kmh - stations_used$all_mean_speed_kmh)^2 +
                                       (stations_used$summer_mean_speed_kmh - stations_used$all_mean_speed_kmh)^2 +
                                       (stations_used$autumn_mean_speed_kmh - stations_used$all_mean_speed_kmh)^2 +
                                       (stations_used$winter_mean_speed_kmh - stations_used$all_mean_speed_kmh)^2)/4)/stations_used$all_mean_speed_kmh

stations_used <- st_as_sf(stations_used,  coords = c("lon", "lat"), crs = 4326, agr = "constant")

library(arcgisbinding)
arc.check_product()
arc.write("./GGRC32_visualization/GGRC32_visualization.gdb/station_speed",
          stations_used, overwrite = TRUE)
```

## How existing bike lines compliments the Toronto Bike Share System?

```{r}
for (i in unique(routes_df$fromId)) {
  trips_starts_at_i <- trips[which(trips$Start.Station.Name == i), ]
  routes_starts_at_i <- routes_df$fromId == i
  for (j in unique(routes_df$toId)) {
    routes_df[routes_starts_at_i & routes_df$toId == j, "n"] <- 
      nrow(trips_starts_at_i[trips_starts_at_i$End.Station.Name == j,])
  }
  cat("process iterated", i, "\n")
}

# store routes that have been used
routes_used_df <- routes_df[routes_df$n > 0,]
```

Adding overlapping routes on top of each other

note: trips start and end at the same station are omitted in the overlay (no route)

```{r}
routes_used <- st_as_sf(routes_used_df, sf_column_name = "geometry")

library(stplanr)
overlay_routes <- overline(sl = routes_used,
                           attrib = "n",
                           ncores = 8)

library(arcgisbinding)
arc.check_product()
arc.write("./GGRC32_visualization/GGRC32_visualization.gdb/overlay_routes1",
          overlay_routes, overwrite = TRUE)
```

## Some descriptive statistics

```{r}
nrow(trips[trips$Start.Station.Name != trips$End.Station.Name,])
nrow(trips_spring[trips_spring$Start.Station.Name != trips_spring$End.Station.Name,])
nrow(trips_summer[trips_summer$Start.Station.Name != trips_summer$End.Station.Name,])
nrow(trips_autumn[trips_autumn$Start.Station.Name != trips_autumn$End.Station.Name,])
nrow(trips_winter[trips_winter$Start.Station.Name != trips_winter$End.Station.Name,])

mean(trips[trips$Start.Station.Name != trips$End.Station.Name, "Trip..Duration"])
median(trips[trips$Start.Station.Name != trips$End.Station.Name, "Trip..Duration"])
IQR(trips[trips$Start.Station.Name != trips$End.Station.Name, "Trip..Duration"])

mean(trips_spring[trips_spring$Start.Station.Name != trips_spring$End.Station.Name, "Trip..Duration"])
median(trips_spring[trips_spring$Start.Station.Name != trips_spring$End.Station.Name, "Trip..Duration"])
IQR(trips_spring[trips_spring$Start.Station.Name != trips_spring$End.Station.Name, "Trip..Duration"])

mean(trips_summer[trips_summer$Start.Station.Name != trips_summer$End.Station.Name, "Trip..Duration"])
median(trips_summer[trips_summer$Start.Station.Name != trips_summer$End.Station.Name, "Trip..Duration"])
IQR(trips_summer[trips_summer$Start.Station.Name != trips_summer$End.Station.Name, "Trip..Duration"])

mean(trips_autumn[trips_autumn$Start.Station.Name != trips_autumn$End.Station.Name, "Trip..Duration"])
median(trips_autumn[trips_autumn$Start.Station.Name != trips_autumn$End.Station.Name, "Trip..Duration"])
IQR(trips_autumn[trips_autumn$Start.Station.Name != trips_autumn$End.Station.Name, "Trip..Duration"])

mean(trips_winter[trips_winter$Start.Station.Name != trips_winter$End.Station.Name, "Trip..Duration"])
median(trips_winter[trips_winter$Start.Station.Name != trips_winter$End.Station.Name, "Trip..Duration"])
IQR(trips_winter[trips_winter$Start.Station.Name != trips_winter$End.Station.Name, "Trip..Duration"])
```

```{r}
nrow(stations_used[!is.nan(stations_used$spring_mean_speed_kmh),])
nrow(stations_used[!is.nan(stations_used$summer_mean_speed_kmh),])
nrow(stations_used[!is.nan(stations_used$autumn_mean_speed_kmh),])
nrow(stations_used[!is.nan(stations_used$winter_mean_speed_kmh),])

mean(stations_used$all_mean_speed_kmh)
median(stations_used$all_mean_speed_kmh)
IQR(stations_used$all_mean_speed_kmh)

mean(stations_used$spring_mean_speed_kmh, na.rm = TRUE)
median(stations_used$spring_mean_speed_kmh, na.rm = TRUE)
IQR(stations_used$spring_mean_speed_kmh, na.rm = TRUE)

mean(stations_used$summer_mean_speed_kmh, na.rm = TRUE)
median(stations_used$summer_mean_speed_kmh, na.rm = TRUE)
IQR(stations_used$summer_mean_speed_kmh, na.rm = TRUE)

mean(stations_used$autumn_mean_speed_kmh, na.rm = TRUE)
median(stations_used$autumn_mean_speed_kmh, na.rm = TRUE)
IQR(stations_used$autumn_mean_speed_kmh, na.rm = TRUE)

mean(stations_used$winter_mean_speed_kmh, na.rm = TRUE)
median(stations_used$winter_mean_speed_kmh, na.rm = TRUE)
IQR(stations_used$winter_mean_speed_kmh, na.rm = TRUE)

mean(stations_used$seasonal_variation, na.rm = TRUE)
median(stations_used$seasonal_variation, na.rm = TRUE)
IQR(stations_used$seasonal_variation, na.rm = TRUE)
```

```{r}
nrow(routes_df)
mean(routes_df$distance)
median(routes_df$distance)
IQR(routes_df$distance)

nrow(routes_used_df)
mean(routes_used_df$distance)
median(routes_used_df$distance)
IQR(routes_used_df$distance)
```

```{r}
nrow(overlay_routes)
mean(overlay_routes$n)
median(overlay_routes$n)
IQR(overlay_routes$n)
```
